### 2015-05-13 whoop.tex
Worked on watchdog race checking.

\medskip\noindent\textbf{Lockset Instrumentation }
%
For thread templates $S$ and $T$ (including the possibility that $S$ and $T$ are equal) we wish to check whether it is possible for a thread executing $S$ to race with a thread executing $T$, in the presence of arbitrarily many further concurrently executing threads.

@@

Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $S$ and $T$ are the same.

@@

\medskip\noindent\textbf{Identifying Set of All Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.

Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.

\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise}). Such sets can be typically modeled in Boogie using maps. However, this requires the use of quantifiers to express invariants related to the contents of a set, such as emptiness. To avoid using quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, which works as follows:

We introduce a global watchdog variable for each SMACK memory region that is common to both entry points of a pair. This variable is an \emph{unconstrained} constant representing the offset to memory region that should be checked for races. Verification involves proving that two entry points cannot race at the watched offset of every memory region. The arbitrary watched offset implies that every offset of each memory region is race-free. Watchdog race-checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.

>>>>

\medskip\noindent\textbf{Lockset Instrumentation }
%
For thread templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we wish to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.

@@

Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $T$ and $U$ are the same.

@@

\medskip\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) we want to advocate the use of good locking practices when developing drivers for the Linux kernel.  \ADComment{Can you re-phrase (ii) to say that using a small, fixed number of locks is advocated by Linux experts as good practice, and cite the drivers book?}

\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.  Suppose we are analysing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $Q$.  When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_Q$, representing some unspecified index into $Q$; we call this the \emph{watched offset} for $Q$.  We then attempt to prove that it is impossible for $T$ and $U$ to race on $Q$ at index $\mathit{watched}_Q$.  If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $Q$, since the watched offset was arbitrary.  This technique of choosing an arbitrary index to analyse for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $Q$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $Q$.
